---
title: "What Drives Poor Mental Health Across U.S. Communities? A County-Level Analysis of Social and Economic Determinants in the United States"
date: "4/15/2026"
format: html
editor: visual
embed-resources: true
---

# Data Processing

## Load and inspect PLACES mental health data

I begin by loading and inspecting the PLACES data so I can confirm its structure, identify the county-level mental health variables available, and isolate the outcome used in the analysis.

```{r}
places <- read.csv("places_county_2025.csv")
```

```{r}
dim(places)
```

```{r}
head(places)
```

```{r}
colnames(places)
```

```{r}
str(places)
```

```{r}
library(dplyr)

places_mh <- places %>%
  select(
    StateAbbr,
    StateDesc,
    CountyName,
    CountyFIPS,
    TotalPopulation,
    TotalPop18plus,
    MHLTH_AdjPrev,
    MHLTH_Adj95CI
  ) %>%
  rename(
    poor_mental_health = MHLTH_AdjPrev,
    poor_mental_health_ci = MHLTH_Adj95CI
  )
```

```{r}
dim(places_mh)
```

```{r}
head(places_mh)
```

```{r}
str(places_mh)
```

## Load and inspect ACS predictors

I next load the ACS data to obtain the county-level social and economic predictors and prepare them for merging with the PLACES mental health outcome.

```{r}
library(tidycensus)

api_key <- Sys.getenv("CENSUS_API_KEY")
census_api_key(api_key, install = FALSE)
```

```{r}
library(tidycensus)
library(dplyr)
library(tidyr)

api_key <- Sys.getenv("CENSUS_API_KEY")
census_api_key(api_key, install = FALSE)

acs_raw <- get_acs(
  geography = "county",
  year = 2023,
  survey = "acs5",
  variables = c(
    poverty_rate = "S1701_C03_001",
    median_household_income = "B19013_001",
    unemployment_rate = "DP03_0009PE",
    bachelors_or_higher = "DP02_0068PE",
    uninsured_rate = "S2701_C05_001",
    rent_burden_30_34_9 = "DP04_0141PE",
    rent_burden_35_plus = "DP04_0142PE"
  )
)
```

```{r}
unique(acs_raw$variable)
```

```{r}
acs_county <- acs_raw %>%
  select(GEOID, NAME, variable, estimate) %>%
  pivot_wider(names_from = variable, values_from = estimate) %>%
  rename(
    bachelors_or_higher = DP02_0068P,
    unemployment_rate = DP03_0009P,
    rent_burden_30_34_9 = DP04_0141P,
    rent_burden_35_plus = DP04_0142P
  ) %>%
  mutate(
    housing_cost_burden = rent_burden_30_34_9 + rent_burden_35_plus
  ) %>%
  select(
    GEOID,
    NAME,
    poverty_rate,
    median_household_income,
    unemployment_rate,
    bachelors_or_higher,
    uninsured_rate,
    rent_burden_30_34_9,
    rent_burden_35_plus,
    housing_cost_burden
  )
```

```{r}
places_mh <- places_mh %>%
  mutate(CountyFIPS = as.character(CountyFIPS))

acs_county <- acs_county %>%
  mutate(GEOID = as.character(GEOID))
```

```{r}
analysis_df <- places_mh %>%
  left_join(acs_county, by = c("CountyFIPS" = "GEOID"))
```

## Load and inspect Census GEOIDs and TIGER/Line shapefiles

Finally, I load county geographic identifiers and shapefiles so the merged dataset can be mapped and explored spatially across U.S. counties.

```{r}
library(tigris)
library(dplyr)

counties_sf <- counties(year = 2024, cb = TRUE) %>%
  mutate(GEOID = as.character(GEOID))

analysis_df <- analysis_df %>%
  mutate(CountyFIPS = as.character(CountyFIPS))

map_df <- counties_sf %>%
  left_join(analysis_df, by = c("GEOID" = "CountyFIPS"))
```

## EDA plots

To begin exploration, I first examine the distribution of the outcome variable. This helps show how poor mental health is distributed across counties and whether there are signs of skewness, clustering, or unusual values.

```{r}
# Outcome distribution
library(ggplot2)
library(dplyr)

ggplot(analysis_df, aes(x = poor_mental_health)) +
  geom_histogram(bins = 30) +
  labs(
    title = "Distribution of Poor Mental Health Across U.S. Counties",
    x = "Age-adjusted prevalence of frequent mental distress",
    y = "Number of counties"
  ) +
  theme_minimal()
```

Next, I examine the distributions of the ACS predictor variables. This helps identify their ranges and variation across counties and can reveal whether any predictors are especially skewed or concentrated.

```{r}
## Predictor distributions
library(tidyr)
library(ggplot2)
library(dplyr)

analysis_df %>%
  select(
    poverty_rate,
    median_household_income,
    unemployment_rate,
    bachelors_or_higher,
    uninsured_rate,
    housing_cost_burden
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = "variable",
    values_to = "value"
  ) %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 30) +
  facet_wrap(~ variable, scales = "free") +
  labs(
    title = "Distributions of ACS Predictor Variables",
    x = "Value",
    y = "Number of counties"
  ) +
  theme_minimal()
```

I then create scatterplots of poor mental health against each predictor. This provides an initial look at the direction and apparent strength of the relationship between the outcome and each county-level factor.

```{r}
## Scatterplots of poor mental health vs each predictor
analysis_df %>%
  select(
    poor_mental_health,
    poverty_rate,
    median_household_income,
    unemployment_rate,
    bachelors_or_higher,
    uninsured_rate,
    housing_cost_burden
  ) %>%
  pivot_longer(
    cols = -poor_mental_health,
    names_to = "predictor",
    values_to = "value"
  ) %>%
  ggplot(aes(x = value, y = poor_mental_health)) +
  geom_point(alpha = 0.4, size = 1) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~ predictor, scales = "free_x") +
  labs(
    title = "Poor Mental Health vs. Social and Economic Predictors",
    x = "Predictor value",
    y = "Poor mental health prevalence"
  ) +
  theme_minimal()
```

Because some predictors appear more strongly related to poor mental health than others, I also look at a few of the clearest bivariate relationships separately. These focused plots make it easier to inspect the patterns for poverty, income, and education in more detail.

```{r}
# A focused set of scatterplots for the strongest-looking predictors
ggplot(analysis_df, aes(x = poverty_rate, y = poor_mental_health)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Poor Mental Health vs. Poverty Rate",
    x = "Poverty rate",
    y = "Poor mental health prevalence"
  ) +
  theme_minimal()
```

```{r}
ggplot(analysis_df, aes(x = median_household_income, y = poor_mental_health)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Poor Mental Health vs. Median Household Income",
    x = "Median household income",
    y = "Poor mental health prevalence"
  ) +
  theme_minimal()
```

```{r}
ggplot(analysis_df, aes(x = bachelors_or_higher, y = poor_mental_health)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Poor Mental Health vs. Bachelor's Degree Attainment",
    x = "Percent with bachelor's degree or higher",
    y = "Poor mental health prevalence"
  ) +
  theme_minimal()
```

To understand how the predictors relate to one another as well as to the outcome, I compute a correlation heatmap. This is useful for identifying which variables move together and whether some predictors may capture overlapping dimensions of disadvantage.

```{r}
# Correlation heatmap
library(corrplot)

cor_df <- analysis_df %>%
  select(
    poor_mental_health,
    poverty_rate,
    median_household_income,
    unemployment_rate,
    bachelors_or_higher,
    uninsured_rate,
    housing_cost_burden
  ) %>%
  na.omit()

cor_matrix <- cor(cor_df)

corrplot(cor_matrix, method = "color", type = "upper", tl.cex = 0.8)
```

I also map county-level poor mental health to explore whether the outcome shows broad geographic patterning. A choropleth is helpful for seeing whether higher burdens appear clustered in particular parts of the country.

```{r}
# County-level choropleth of poor mental health
library(ggplot2)
library(dplyr)

ggplot(map_df) +
  geom_sf(aes(fill = poor_mental_health), color = NA) +
  coord_sf(
    xlim = c(-125, -66),
    ylim = c(24, 50),
    expand = FALSE
  ) +
  labs(
    title = "Poor Mental Health Across U.S. Counties",
    fill = "Poor mental\nhealth"
  ) +
  theme_minimal()
```

Finally, I compare average poor mental health across states as a simple geographic summary. This provides a quick way to see whether some states tend to have systematically higher county-level poor mental health than others.

```{r}
# Quick regional-style comparison using state averages
analysis_df %>%
  group_by(StateAbbr) %>%
  summarize(mean_poor_mental_health = mean(poor_mental_health, na.rm = TRUE)) %>%
  ggplot(aes(x = reorder(StateAbbr, mean_poor_mental_health), y = mean_poor_mental_health)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Average Poor Mental Health by State",
    x = "State",
    y = "Average poor mental health prevalence"
  ) +
  theme_minimal()
```