---
title: "What Drives Poor Mental Health Across U.S. Communities? A County-Level Analysis of Social and Economic Determinants in the United States"
date: "4/15/2026"
author: "Andre Gala-Garza"
format: html
editor: visual
embed-resources: true
---

**Disclaimer:** Generative AI was used to assist with drafting report information and templating and writing code in this assignment; however, this writing was checked manually and edited by hand to ensure accuracy.

**Source:** OpenAI. (2026). *ChatGPT* \[Large language model\]. <https://chatgpt.com/>.

# Introduction

Mental health is a core part of overall well-being, and it is shaped not only by individual circumstances but also by the broader conditions in which people live. The CDC notes that mental health is influenced by factors at the individual, family, community, and societal levels, which aligns with the social determinants of health framework (Centers for Disease Control and Prevention, "About Mental Health"; Office of Disease Prevention and Health Promotion). Because these conditions affect stress, stability, access to care, and quality of life, they may help explain why some communities experience a higher burden of poor mental health than others.

This project asks: **Which social and economic factors are most strongly associated with poor mental health across U.S. counties?** It also asks a follow-up question: **Do the counties with the worst mental health outcomes simply reflect higher structural disadvantage, or do they appear to have worse mental health than expected even after accounting for those factors?** A county-level approach makes it possible to study broad geographic patterns across differences in poverty, education, insurance coverage, and housing strain.

The outcome variable comes from the CDC’s PLACES dataset, which provides model-based local health estimates across the United States. I use the county-level measure Frequent Mental Distress among adults, defined as the percentage of adults reporting that their mental health was "not good" for 14 or more of the past 30 days (Centers for Disease Control and Prevention, "Health Status \| PLACES"). Because this project compares counties, I use the age-adjusted prevalence of this measure.

To examine possible correlates of poor mental health, I merge PLACES data with county-level indicators from the American Community Survey (ACS) 5-year estimates, including poverty, household income, unemployment, educational attainment, insurance coverage, and housing conditions (United States Census Bureau). Together, these measures provide a county-level dataset for assessing how structural social and economic conditions are associated with poor mental health across the United States.

This project does not aim to make causal claims; instead, it aims to identify which county-level conditions are most closely associated with poor mental health and whether the highest-burden counties appear worse than those conditions alone would predict.

# Methods

To answer this research question, I combined county-level data from two public sources. The outcome variable comes from the CDC’s 2025 PLACES county dataset, from which I use the age-adjusted prevalence of frequent mental distress among adults (MHLTH_AdjPrev) as my measure of poor mental health. To examine possible social and economic drivers of this outcome, I merged the PLACES data with county-level estimates from the American Community Survey (ACS) 5-year data obtained through the Census API.

The ACS predictors in this analysis are poverty rate, median household income, unemployment rate, percentage of adults with a bachelor’s degree or higher, uninsured rate, and housing cost burden. These variables capture several major dimensions of community conditions, including economic hardship, labor market stability, educational attainment, access to health insurance, and housing-related financial strain. By joining the PLACES and ACS datasets using county FIPS/GEOID identifiers, I created a county-level dataset to examine how structural social and economic factors are associated with poor mental health across U.S. communities.

# Results

## County-Level Mental Health Overview

I first present a publication-ready table showing counties with the highest levels of poor mental health. This helps identify which counties appear at the top of the distribution before turning to a national map.

```{r}
#| echo: false
#| include: false
#| message: false
#| warning: false

library(dplyr)
library(tidyr)
library(tidycensus)
library(tigris)

places <- read.csv("places_county_2025.csv")

places_mh <- places %>%
  select(
    StateAbbr,
    StateDesc,
    CountyName,
    CountyFIPS,
    TotalPopulation,
    TotalPop18plus,
    MHLTH_AdjPrev,
    MHLTH_Adj95CI
  ) %>%
  rename(
    poor_mental_health = MHLTH_AdjPrev,
    poor_mental_health_ci = MHLTH_Adj95CI
  ) %>%
  mutate(CountyFIPS = as.character(CountyFIPS))

api_key <- Sys.getenv("CENSUS_API_KEY")
census_api_key(api_key, install = FALSE)

acs_raw <- get_acs(
  geography = "county",
  year = 2023,
  survey = "acs5",
  variables = c(
    poverty_rate = "S1701_C03_001",
    median_household_income = "B19013_001",
    unemployment_rate = "DP03_0009PE",
    bachelors_or_higher = "DP02_0068PE",
    uninsured_rate = "S2701_C05_001",
    rent_burden_30_34_9 = "DP04_0141PE",
    rent_burden_35_plus = "DP04_0142PE"
  )
)

acs_county <- acs_raw %>%
  select(GEOID, NAME, variable, estimate) %>%
  pivot_wider(names_from = variable, values_from = estimate) %>%
  rename(
    bachelors_or_higher = DP02_0068P,
    unemployment_rate = DP03_0009P,
    rent_burden_30_34_9 = DP04_0141P,
    rent_burden_35_plus = DP04_0142P
  ) %>%
  mutate(
    housing_cost_burden = rent_burden_30_34_9 + rent_burden_35_plus,
    GEOID = as.character(GEOID)
  ) %>%
  select(
    GEOID,
    NAME,
    poverty_rate,
    median_household_income,
    unemployment_rate,
    bachelors_or_higher,
    uninsured_rate,
    housing_cost_burden
  )

analysis_df <- places_mh %>%
  left_join(acs_county, by = c("CountyFIPS" = "GEOID"))

counties_sf <- counties(year = 2024, cb = TRUE) %>%
  mutate(GEOID = as.character(GEOID))

analysis_df <- analysis_df %>%
  mutate(CountyFIPS = as.character(CountyFIPS))

map_df <- counties_sf %>%
  left_join(analysis_df, by = c("GEOID" = "CountyFIPS"))
```

```{r}
#| label: tbl-mental-health-county
#| echo: false
#| message: false
#| warning: false

library(dplyr)
library(knitr)
library(kableExtra)
library(scales)

mental_health_table <- analysis_df %>%
  select(StateDesc, CountyName, poor_mental_health, TotalPopulation) %>%
  mutate(
    TotalPopulation = as.numeric(gsub(",", "", TotalPopulation))
  ) %>%
  arrange(desc(poor_mental_health)) %>%
  slice_head(n = 15) %>%
  mutate(
    poor_mental_health = round(poor_mental_health, 1),
    TotalPopulation = comma(TotalPopulation)
  ) %>%
  rename(
    State = StateDesc,
    County = CountyName,
    `Poor Mental Health (%)` = poor_mental_health,
    Population = TotalPopulation
  )

mental_health_table %>%
  kable(
    caption = "Top 15 U.S. Counties by Age-Adjusted Prevalence of Frequent Mental Distress",
    align = c("l", "l", "r", "r"),
    booktabs = TRUE
  ) %>%
  kable_styling(
    full_width = FALSE,
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    position = "center"
  ) %>%
  row_spec(0, bold = TRUE)
```

The table shows that the counties with the highest prevalence of poor mental health are heavily concentrated in West Virginia, with a few additional counties from Tennessee and South Dakota. The percentages are fairly tightly clustered, ranging from about 24.4% to 26.6%, which suggests that the highest-burden counties are all experiencing similarly elevated levels of frequent mental distress.

Next, I map county-level poor mental health across the United States using an interactive choropleth map. This visualization makes it easier to see broad geographic patterns and identify regions where poor mental health appears especially high or low.

```{r}
#| label: fig-mental-health-map
#| echo: false
#| message: false
#| warning: false

library(leaflet)
library(sf)
library(dplyr)
library(htmltools)
library(scales)

map_data <- map_df %>%
  st_transform(4326) %>%
  mutate(
    popup_text = paste0(
      "<strong>", CountyName, ", ", StateAbbr, "</strong><br>",
      "Poor mental health: ", round(poor_mental_health, 1), "%<br>",
      "Poverty rate: ", round(poverty_rate, 1), "%<br>",
      "Median household income: ", dollar(median_household_income), "<br>",
      "Unemployment rate: ", round(unemployment_rate, 1), "%<br>",
      "Bachelor's degree or higher: ", round(bachelors_or_higher, 1), "%<br>",
      "Uninsured rate: ", round(uninsured_rate, 1), "%<br>",
      "Housing cost burden: ", round(housing_cost_burden, 1), "%"
    )
  )

pal <- colorNumeric(
  palette = "YlOrRd",
  domain = map_data$poor_mental_health,
  na.color = "#d9d9d9"
)

leaflet(map_data) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  fitBounds(
    lng1 = -125,
    lat1 = 24,
    lng2 = -66,
    lat2 = 50
  ) %>%
  addPolygons(
    fillColor = ~pal(poor_mental_health),
    color = "white",
    weight = 0.3,
    fillOpacity = 0.8,
    smoothFactor = 0.2,
    popup = ~popup_text,
    highlightOptions = highlightOptions(
      weight = 1.2,
      color = "#333333",
      fillOpacity = 0.9,
      bringToFront = TRUE
    )
  ) %>%
  addLegend(
    pal = pal,
    values = ~poor_mental_health,
    opacity = 0.8,
    title = "Poor mental health (%)",
    position = "bottomright"
  )
```

The map suggests broad spatial clustering, with relatively high levels of poor mental health concentrated in parts of Appalachia, the Southeast, and portions of the rural South, while lower levels appear more common in parts of the Great Plains, Upper Midwest, and some Northeastern counties. Overall, the choropleth indicates that poor mental health is not randomly distributed, but instead shows clear regional patterns across U.S. counties.

## Associations Between Poor Mental Health and County-Level Predictors

To explore how poor mental health varies with each county-level predictor, I next create a set of side-by-side scatterplots. Faceting the plots allows for direct visual comparison of the direction and strength of the relationship between poor mental health and each predictor.

```{r}
#| label: fig-scatter-predictors
#| echo: false
#| message: false
#| warning: false
#| fig-width: 12
#| fig-height: 8
#| fig-cap: "Scatterplots of poor mental health versus county-level social and economic predictors."

library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)

scatter_df <- analysis_df %>%
  select(
    poor_mental_health,
    poverty_rate,
    median_household_income,
    unemployment_rate,
    bachelors_or_higher,
    uninsured_rate,
    housing_cost_burden
  ) %>%
  pivot_longer(
    cols = -poor_mental_health,
    names_to = "predictor",
    values_to = "value"
  ) %>%
  mutate(
    predictor = factor(
      predictor,
      levels = c(
        "poverty_rate",
        "median_household_income",
        "unemployment_rate",
        "bachelors_or_higher",
        "uninsured_rate",
        "housing_cost_burden"
      ),
      labels = c(
        "Poverty rate (%)",
        "Median household income ($)",
        "Unemployment rate (%)",
        "Bachelor's degree or higher (%)",
        "Uninsured rate (%)",
        "Housing cost burden (%)"
      )
    )
  )

rainbow_colors <- c(
  "Poverty rate (%)" = "#F4A3A3",
  "Median household income ($)" = "#F7C58B",
  "Unemployment rate (%)" = "#F6E58D",
  "Bachelor's degree or higher (%)" = "#B8E6B8",
  "Uninsured rate (%)" = "#A9D6F5",
  "Housing cost burden (%)" = "#D6B3E6"
)

ggplot(scatter_df, aes(x = value, y = poor_mental_health, color = predictor)) +
  geom_point(alpha = 0.35, size = 1.2) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 0.8) +
  facet_wrap(~ predictor, scales = "free_x", ncol = 2) +
  scale_color_manual(values = rainbow_colors) +
  labs(
    x = NULL,
    y = "Poor mental health (%)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    strip.text = element_text(face = "bold"),
    axis.title.y = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold"),
    legend.position = "none"
  )
```

The scatterplots suggest that counties with higher poverty, unemployment, uninsured rates, and housing cost burden tend to have higher levels of poor mental health, while counties with higher median household income and higher educational attainment tend to have lower levels of poor mental health. The clearest patterns appear for poverty rate, median household income, unemployment rate, and bachelor’s degree attainment, while the relationship with housing cost burden looks weaker and more diffuse.

## Regional Differences in Poor Mental Health

To examine whether poor mental health differs across broad parts of the country, I group counties into U.S. Census regions and compare their distributions using boxplots. This helps show whether some regions tend to have higher or lower county-level burdens of poor mental health, beyond what is visible in the national map.

```{r}
#| label: fig-region-boxplot
#| echo: false
#| message: false
#| warning: false
#| fig-width: 10
#| fig-height: 6
#| fig-cap: "Distribution of poor mental health across U.S. Census regions."

library(dplyr)
library(ggplot2)
library(forcats)

region_df <- analysis_df %>%
  mutate(
    region = case_when(
      StateAbbr %in% c("CT", "ME", "MA", "NH", "RI", "VT",
                       "NJ", "NY", "PA") ~ "Northeast",
      StateAbbr %in% c("IL", "IN", "MI", "OH", "WI",
                       "IA", "KS", "MN", "MO", "NE", "ND", "SD") ~ "Midwest",
      StateAbbr %in% c("DE", "FL", "GA", "MD", "NC", "SC", "VA", "DC", "WV",
                       "AL", "KY", "MS", "TN", "AR", "LA", "OK", "TX") ~ "South",
      StateAbbr %in% c("AZ", "CO", "ID", "MT", "NV", "NM", "UT", "WY",
                       "AK", "CA", "HI", "OR", "WA") ~ "West",
      TRUE ~ NA_character_
    ),
    region = factor(region, levels = c("Northeast", "Midwest", "South", "West"))
  ) %>%
  filter(!is.na(region), !is.na(poor_mental_health))

region_colors <- c(
  "Northeast" = "#A9D6F5",
  "Midwest" = "#B8E6B8",
  "South" = "#F4A3A3",
  "West" = "#D6B3E6"
)

ggplot(region_df, aes(x = region, y = poor_mental_health, fill = region)) +
  geom_boxplot(color = "gray30", outlier.alpha = 0.3) +
  scale_fill_manual(values = region_colors) +
  labs(
    x = "Region",
    y = "Poor mental health (%)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.title = element_text(face = "bold"),
    axis.text.x = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    legend.position = "none"
  )
```

The boxplots show that counties in the South tend to have the highest median levels of poor mental health and also a somewhat wider spread than the other regions. The Northeast appears to have the lowest median, while the Midwest and West fall in between, suggesting meaningful regional differences in county-level mental health burden.

## How Do the Highest-Burden Counties Compare to the National Pattern?

To connect the focused analysis of the highest-burden counties back to the full national sample, I compare each county’s observed poor mental health to the value predicted by the regression model. Counties above the 45-degree line have worse mental health than expected given their structural disadvantage, while counties below the line have better mental health than expected.

```{r}
#| label: fig-observed-vs-predicted
#| echo: false
#| message: false
#| warning: false
#| fig-width: 10
#| fig-height: 7
#| fig-cap: "Observed versus predicted poor mental health for all U.S. counties, with the top 15 highest-burden counties highlighted."

library(dplyr)
library(ggplot2)
library(ggrepel)

obs_pred_df <- analysis_df %>%
  select(
    CountyFIPS,
    CountyName,
    StateAbbr,
    poor_mental_health,
    poverty_rate,
    median_household_income,
    unemployment_rate,
    bachelors_or_higher,
    uninsured_rate,
    housing_cost_burden
  ) %>%
  na.omit()

obs_pred_model <- lm(
  poor_mental_health ~ poverty_rate +
    median_household_income +
    unemployment_rate +
    bachelors_or_higher +
    uninsured_rate +
    housing_cost_burden,
  data = obs_pred_df
)

obs_pred_df <- obs_pred_df %>%
  mutate(
    predicted_mental_health = predict(obs_pred_model, newdata = obs_pred_df)
  )

top_15_counties <- obs_pred_df %>%
  arrange(desc(poor_mental_health)) %>%
  slice_head(n = 15) %>%
  mutate(
    county_label = paste0(CountyName, ", ", StateAbbr)
  ) %>%
  select(CountyFIPS, county_label)

obs_pred_df <- obs_pred_df %>%
  left_join(top_15_counties, by = "CountyFIPS") %>%
  mutate(
    group = ifelse(is.na(county_label), "All other counties", "Top 15 counties")
  )

plot_limits <- range(
  c(obs_pred_df$predicted_mental_health, obs_pred_df$poor_mental_health),
  na.rm = TRUE
)

ggplot(obs_pred_df, aes(x = predicted_mental_health, y = poor_mental_health)) +
  geom_abline(
    slope = 1,
    intercept = 0,
    linetype = "dashed",
    color = "gray50",
    linewidth = 0.8
  ) +
  geom_point(
    data = filter(obs_pred_df, group == "All other counties"),
    color = "gray70",
    alpha = 0.5,
    size = 1.5
  ) +
  geom_point(
    data = filter(obs_pred_df, group == "Top 15 counties"),
    color = "#B22222",
    size = 2.5
  ) +
  geom_text_repel(
    data = filter(obs_pred_df, group == "Top 15 counties"),
    aes(label = county_label),
    size = 3.2,
    box.padding = 0.3,
    point.padding = 0.2,
    max.overlaps = 20
  ) +
  coord_equal(xlim = plot_limits, ylim = plot_limits) +
  labs(
    x = "Predicted poor mental health (%)",
    y = "Observed poor mental health (%)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.title = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    legend.position = "none"
  )
```

The highlighted counties fall mostly above the 45-degree line, indicating that their observed poor mental health is generally higher than what the regression model predicts based on structural disadvantage alone. This suggests that while factors like poverty, income, education, unemployment, insurance, and housing burden explain part of their high mental health burden, many of these counties—especially in West Virginia—still appear unusually unhealthy even relative to similar counties across the United States.

## How Do the Most Unhealthy Counties Differ from U.S. Counties Overall?

To compare the most unhealthy counties to the national county distribution, I standardize each predictor across all counties and then calculate the average standardized value for the top 15 counties and for all U.S. counties. I display those averages in a dumbbell plot so the gap between the two groups is easier to compare across predictors on a common scale.

```{r}
#| label: fig-top15-vs-us-dumbbell-standardized
#| echo: false
#| message: false
#| warning: false
#| fig-width: 10
#| fig-height: 6
#| fig-cap: "Dumbbell plot comparing standardized predictor averages for the top 15 highest-burden counties and all U.S. counties."

library(dplyr)
library(tidyr)
library(ggplot2)
library(forcats)

comparison_df <- analysis_df %>%
  select(
    CountyFIPS,
    CountyName,
    StateAbbr,
    poor_mental_health,
    poverty_rate,
    median_household_income,
    unemployment_rate,
    bachelors_or_higher,
    uninsured_rate,
    housing_cost_burden
  ) %>%
  na.omit() %>%
  mutate(
    poverty_rate = as.numeric(scale(poverty_rate)),
    median_household_income = as.numeric(scale(median_household_income)),
    unemployment_rate = as.numeric(scale(unemployment_rate)),
    bachelors_or_higher = as.numeric(scale(bachelors_or_higher)),
    uninsured_rate = as.numeric(scale(uninsured_rate)),
    housing_cost_burden = as.numeric(scale(housing_cost_burden))
  )

top_15_ids <- comparison_df %>%
  arrange(desc(poor_mental_health)) %>%
  slice_head(n = 15) %>%
  pull(CountyFIPS)

group_means <- comparison_df %>%
  mutate(
    group = if_else(CountyFIPS %in% top_15_ids, "Top 15 counties", "All U.S. counties")
  ) %>%
  group_by(group) %>%
  summarize(
    poverty_rate = mean(poverty_rate, na.rm = TRUE),
    median_household_income = mean(median_household_income, na.rm = TRUE),
    unemployment_rate = mean(unemployment_rate, na.rm = TRUE),
    bachelors_or_higher = mean(bachelors_or_higher, na.rm = TRUE),
    uninsured_rate = mean(uninsured_rate, na.rm = TRUE),
    housing_cost_burden = mean(housing_cost_burden, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_longer(
    cols = -group,
    names_to = "predictor",
    values_to = "mean_value"
  ) %>%
  mutate(
    predictor = recode(
      predictor,
      poverty_rate = "Poverty rate",
      median_household_income = "Median household income",
      unemployment_rate = "Unemployment rate",
      bachelors_or_higher = "Bachelor's degree or higher",
      uninsured_rate = "Uninsured rate",
      housing_cost_burden = "Housing cost burden"
    )
  ) %>%
  pivot_wider(names_from = group, values_from = mean_value) %>%
  mutate(
    gap = `Top 15 counties` - `All U.S. counties`,
    predictor = fct_reorder(predictor, gap)
  )

points_df <- group_means %>%
  pivot_longer(
    cols = c(`All U.S. counties`, `Top 15 counties`),
    names_to = "group",
    values_to = "mean_value"
  )

ggplot(group_means, aes(y = predictor)) +
  geom_segment(
    aes(
      x = `All U.S. counties`,
      xend = `Top 15 counties`,
      yend = predictor
    ),
    color = "gray70",
    linewidth = 1
  ) +
  geom_point(
    data = points_df,
    aes(x = mean_value, shape = group, color = group),
    size = 3.4
  ) +
  scale_shape_manual(
    values = c("All U.S. counties" = 16, "Top 15 counties" = 17)
  ) +
  scale_color_manual(
    values = c("All U.S. counties" = "#2166AC", "Top 15 counties" = "#B22222")
  ) +
  labs(
    x = "Average standardized value (z-score)",
    y = NULL,
    shape = NULL,
    color = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.title.x = element_text(face = "bold"),
    axis.text.y = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank(),
    legend.position = "top"
  )
```

## Which Factors Are Most Strongly Associated with Poor Mental Health Across U.S. Counties?

To directly answer the main research question, I fit a multiple linear regression model using all six county-level predictors and compare their standardized coefficients. I display the results as a diverging lollipop plot so it is easier to compare the relative strength and direction of each association on a common scale.

```{r}
#| label: fig-final-diverging-lollipop
#| echo: false
#| message: false
#| warning: false
#| fig-width: 10
#| fig-height: 6
#| fig-cap: "Diverging lollipop plot of standardized coefficients from a multiple linear regression model predicting poor mental health across U.S. counties."

library(dplyr)
library(ggplot2)
library(broom)
library(forcats)

final_reg_df <- analysis_df %>%
  select(
    poor_mental_health,
    poverty_rate,
    median_household_income,
    unemployment_rate,
    bachelors_or_higher,
    uninsured_rate,
    housing_cost_burden
  ) %>%
  na.omit() %>%
  mutate(across(everything(), ~ as.numeric(scale(.x))))

final_model <- lm(
  poor_mental_health ~ poverty_rate +
    median_household_income +
    unemployment_rate +
    bachelors_or_higher +
    uninsured_rate +
    housing_cost_burden,
  data = final_reg_df
)

final_coef_df <- tidy(final_model, conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    term = recode(
      term,
      poverty_rate = "Poverty rate",
      median_household_income = "Median household income",
      unemployment_rate = "Unemployment rate",
      bachelors_or_higher = "Bachelor's degree or higher",
      uninsured_rate = "Uninsured rate",
      housing_cost_burden = "Housing cost burden"
    ),
    direction = if_else(estimate > 0, "Positive", "Negative"),
    term = fct_reorder(term, estimate)
  )

ggplot(final_coef_df, aes(x = estimate, y = term, color = direction)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray55", linewidth = 0.7) +
  geom_segment(
    aes(x = 0, xend = estimate, y = term, yend = term),
    linewidth = 1.1,
    show.legend = FALSE
  ) +
  geom_errorbarh(
    aes(xmin = conf.low, xmax = conf.high),
    height = 0.14,
    color = "gray45",
    linewidth = 0.7
  ) +
  geom_point(size = 3.8, show.legend = FALSE) +
  scale_color_manual(
    values = c("Positive" = "#B22222", "Negative" = "#2166AC")
  ) +
  labs(
    x = "Standardized coefficient",
    y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.title.x = element_text(face = "bold"),
    axis.text.y = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank()
  )
```

The diverging lollipop plot shows that poverty rate has the strongest positive association with poor mental health, while bachelor’s degree attainment has the strongest negative association. Unemployment rate and housing cost burden also have meaningful positive associations, while median household income and uninsured rate are negatively associated, indicating that economic hardship and lower educational attainment are the factors most strongly linked to poor mental health across U.S. counties.

# Conclusion and Summary

This project examined which social and economic factors are most strongly associated with poor mental health across U.S. counties. Poor mental health showed clear geographic clustering, with especially high burdens in parts of Appalachia and the South, and many of the highest-burden counties were located in West Virginia.

Across the analysis, counties with higher poverty, unemployment, and housing cost burden tended to have worse mental health outcomes, while counties with higher income and higher educational attainment tended to have better outcomes. The final standardized coefficient plot suggested that poverty rate had the strongest positive association with poor mental health, while bachelor’s degree attainment had the strongest negative association. Overall, these results indicate that economic hardship and lower educational attainment are the factors most strongly linked to poor mental health in this county-level analysis.

At the same time, several of the highest-burden counties still appeared to have worse mental health than predicted by these structural factors alone, suggesting that the model does not capture everything driving county-level differences. Future research could examine additional county-level factors, such as access to mental health care, rural isolation, or social support, to better explain why some counties experience especially high burdens of poor mental health.

# Works Cited

Centers for Disease Control and Prevention. “About Mental Health.” *CDC*, 9 June 2025, www.cdc.gov/mental-health/about/index.html.

Centers for Disease Control and Prevention. “Health Status \| PLACES.” *CDC*, 29 Oct. 2024, www.cdc.gov/places/measure-definitions/health-status.html.

Office of Disease Prevention and Health Promotion. “Social Determinants of Health.” *Healthy People 2030*, U.S. Department of Health and Human Services, odphp.health.gov/healthypeople/priority-areas/social-determinants-health.

United States Census Bureau. “American Community Survey 5-Year Data (2009–2024).” *U.S. Census Bureau*, www.census.gov/data/developers/data-sets/acs-5year.html.
